---
title: " Oracle soils Burkina Faso t = 0 "
author: "Frédéric Mahé"
date: '`r format(Sys.time(), "%d %B %Y")`'

output:
  rmarkdown::html_document:
    theme: lumen
    toc: yes
    toc_float: TRUE
    keep_md: yes
    # code_folding: hide
---

```{r setup, include=FALSE}
rm(list = ls()) # remove all the object before starting
knitr::opts_chunk$set(echo = TRUE)
```

***

#### load required packages

```{r packages, message=FALSE}
library(here)
library(tidyverse)
```


# SSU rRNA metabarcoding (16S V3-V4)

## decontamination and subsampling

### variables and functions

```{r}
input <- "Oracle_soils_16S_341F_785R_87_samples.OTU.filtered.cleaved.nosubstringOTUs.mumu.table2"
title <- "Oracle MetaB t0 SSU assignment with Silva v138.1 (87 samples)"
rarefied_table <- str_replace(input, "table2", "table3")
useful_metadata <- c("OTU", "total", "amplicon", "length", "abundance",
                   "spread", "identity", "taxonomy")
useless_metadata <- c("sequence", "quality", "cloud", "references", "chimera")
min_target_size <- 10000  # minimal number of reads
problematic_sample <- "T-1_S82"
```

Fix an automatic column type detection issue for sample
`r problematic_sample`:

```{r}
load_raw_16S_occurrence_data <- function(filename){
    here::here("data/MetaB_t0/", filename) %>%
        read_tsv(na = "0",
                 show_col_types = FALSE,
                 col_type = list(`T-1_S82` = col_double()))
}

. %>%
    select(OTU, starts_with("T-"), starts_with("OR-")) %>%
    pivot_longer(cols = -OTU,
                 names_to = "samples",
                 values_to = "reads") %>%
    filter(! is.na(reads)) %>%
    summarise(sum(reads)) %>%
    pull() -> count_all_reads
```


### preliminary taxonomic filtering

Remove unassigned entries and entries assigned to organelles:

```{r}
load_raw_16S_occurrence_data(input) %>%
    filter(! str_detect(taxonomy, "No_hit|Mitochondria|Chloroplast")) -> starting_table
```

How many reads were removed?

```{r}
load_raw_16S_occurrence_data(input) %>%
    count_all_reads -> reads_before

starting_table %>%
    count_all_reads -> reads_after
```

`r prettyNum(reads_before - reads_after, scientific=FALSE, big.mark=",")`
reads were removed. That's a 
`r prettyNum(100 * (1 - (reads_after / reads_before)), scientific=FALSE)`
% reduction.


clean up:

```{r}
rm(problematic_sample, reads_before, reads_after,
   load_raw_16S_occurrence_data, count_all_reads,
   input)
```


### extract control samples

```{r}
## Extract control samples
starting_table %>%
    select(all_of(useful_metadata),
           starts_with("T-"),
           starts_with("OR-T-extr")) %>%
    pivot_longer(cols = -all_of(useful_metadata),
                 names_to = "samples",
                 values_to = "reads") %>%
    filter(! is.na(reads)) %>%
    count(OTU, wt = reads) -> controls

controls
```


### decontamination

Compute the total number of reads for each OTU present in control
samples. That sum is then substracted from all occurrences of that OTU
in true samples. The rationale is as follows:

- if an OTU is abundant in control samples, but rare in true samples,
  then it is a contamination specific to the control samples, and it
  will be eliminated by the substraction (i.e, final abundance is 0),
- if an OTU is present in control samples, and present in true samples
  (systematic contamination, will be mitigated by the substraction),
 - if an OTU is rare in control samples, but abundant in true samples
  (cross-talk, will be eliminated/mitigated by the substraction)

Control samples can be eliminated from the statistical analysis after
the substraction (all OTUs present in control samples have been zeroed
out).


```{r}
starting_table %>%
    select(-all_of(useless_metadata)) %>%
    pivot_longer(cols = -all_of(useful_metadata),
                 names_to = "samples",
                 values_to = "reads") %>%
    filter(! is.na(reads)) %>%
    left_join(controls, by = "OTU") %>%
    mutate(reads = case_when(
               is.na(n)  ~ reads,
               n > reads ~ 0,
               TRUE      ~ reads - n)) %>%
    select(-n) %>%
    pivot_wider(names_from = samples,
                values_from = reads,
                values_fill = 0) -> d

rm(controls, starting_table)
```

Find sample sizes (number of reads per sample):

```{r}
d %>%
    select(-all_of(useful_metadata)) %>%
    pivot_longer(cols = everything(),
                 names_to = "samples",
                 values_to = "reads") %>%
    count(samples, wt = reads, name = "reads") %>%
    arrange(reads) -> sample_sizes
```

Find duplicated samples:

```{r}
sample_sizes %>%
    filter(str_detect(samples, "-S_")) %>%
    mutate(samples = str_remove(samples, "-S_.*")) %>%
    pull(samples) -> duplicated_samples

duplicated_samples
```

How many reads in these duplicated samples?

```{r}
sample_sizes %>%
    pivot_wider(names_from = samples,
                values_from = reads) %>%
    select(starts_with(duplicated_samples)) %>%
    pivot_longer(cols = everything(),
                 names_to = "samples",
                 values_to = "reads")
```

It seems that re-sequenced samples have more reads in all cases.

List small samples that will be eliminated:

```{r}
sample_sizes %>%
    filter(reads < min_target_size) %>%
    select(samples) %>%
    pull() -> small_samples

small_samples
```

Get the size of the smallest remaining samples:

```{r}
sample_sizes %>%
    filter(reads >= min_target_size) %>%
    summarize(min(reads)) %>%
    pull() -> smallest_sample_size
```

The smallest sample contains 
`r prettyNum(smallest_sample_size, scientific=FALSE, big.mark=",")` 
reads.

Isolate metadata columns, rename samples  and prepare rarefaction:

```{r}
d %>%
    select(all_of(useful_metadata)) -> metadata

. %>%
    pivot_longer(cols = -OTU,
                 names_to = "samples",
                 values_to = "reads") %>%
    mutate(samples = str_remove(samples, "-S_"),
           samples = str_replace(samples, "^OR-", "BU-"),
           samples = str_remove(samples, "_?S[:digit:]+$")) %>%
    pivot_wider(names_from = samples,
                values_from = reads) %>%
        select(-OTU) -> rename_samples

d %>%
    select(-all_of(c(useful_metadata, small_samples)), OTU) %>%
    rename_samples %>%
    t() -> occurrences_t
```

To be continued...


***

```{r}
sessionInfo()
```
