---
title: " Oracle soils Burkina Faso t = 0 (#6) "
author: "Frédéric Mahé"
date: '`r format(Sys.time(), "%d %B %Y")`'

output:
  rmarkdown::html_document:
    theme: lumen
    toc: yes
    toc_float: TRUE
    keep_md: yes
    # code_folding: hide
---

```{r setup, include=FALSE}
rm(list = ls()) # remove all the object before starting
knitr::opts_chunk$set(echo = TRUE)
```

***

#### load required packages

```{r packages, message=FALSE}
library(here)
library(tidyverse)
library(vegan)
## library(hilldiv)
```


# SSU rRNA metabarcoding (16S V3-V4)

## preamble

### variables and functions

```{r}
input <- "Oracle_soils_16S_341F_785R_87_samples.OTU.filtered.cleaved.nosubstringOTUs.mumu.table3"
title <- "Oracle MetaB t0 SSU NMDS (80 samples)"
useful_metadata <- c("OTU", "total", "amplicon", "length", "abundance",
                     "spread", "identity", "taxonomy", "references")
cbPalette <- c("#E69F00", "#56B4E9", "#009E73", "#0072B2", "#D55E00", "#CC79A7")
seed <- 123
```

```{r}
load_clean_16S_occurrence_data <- function(filename){
    here::here("data", "MetaB_t0", filename) %>%
        read_tsv(show_col_types = FALSE)
}
```

### load data

```{r}
load_clean_16S_occurrence_data(input) -> full_occurrence_table

```


## alpha diversity

### estimate usual diversity index

extract only occurrences and transpose for vegan

```{r}
full_occurrence_table %>%
    select(-all_of(useful_metadata)) %>%
    t() -> occurrences_t
```

compute indices

```{r}
## richness
vegan::estimateR(round(occurrences_t)) -> R

## Shannon index H; richness + evenness
vegan::diversity(occurrences_t, index = "shannon") -> H

## Simpson's D index; richness + evenness; 0-1; 1 - D rises as evenness increases
vegan::diversity(occurrences_t, index = "simpson") -> D
vegan::diversity(occurrences_t, index = "invsimpson") -> inv_D

## Pielou’s index of evenness; 0-1, 1 = max. evenness
vegan::specnumber(occurrences_t) -> S
H / log(S) -> J
```

### Hill numbers: estimate effective number of species

This equation has a parameter q that defines its sensitivity to rare species:
low values of q favor rare species, high values of q favor abundant species. 

For example, Shannon diversity is of order q = 1, and for Simpson
diversity q = 2.

When q = 0, diversity = S (richness), because rare species are treated
the same as abundant ones.

```{r}
##{{richness as hill numbers}}
## occurrences_t

## dR <- bact_OTUs_rarefied_t %>% 
##   t() %>% 
##   hill_div(qvalue = 0) %>%
##   as_tibble(rownames = NA) %>%
##   rownames_to_column(var = "soil_code")

## #{{richness + evenness (= shannon diversity) as hill numbers; rare species}}
## dREr <- bact_OTUs_rarefied_t %>% 
##   t() %>%
##   hill_div(qvalue = 1) %>%
##   as_tibble(rownames = NA) %>%
##   rownames_to_column(var = "soil_code")

## ##{{richness + evenness (= inverse Simpson = simpson diversity) as hill numbers; abundant species}}
## dREa <- bact_OTUs_rarefied_t %>% 
##   t() %>%
##   hill_div(qvalue = 2) %>%
##   as_tibble(rownames = NA) %>%
##   rownames_to_column(var = "soil_code")

#{{evenness (= shannon evenness) as hill ratio}}
## eDRr <- dREr %>%
##   left_join(dR, by = "soil_code") %>%
##   mutate(value = value.x / value.y) %>%
##   select(soil_code,value)

#{{evenness (= ssimpson evenness) as hill ratio}}
## eDRa  <- dREa %>%
##   left_join(dR, by = "soil_code") %>%
##   mutate(value = value.x / value.y) %>%
##   select(soil_code,value)
```


### pool diversity indices

```{r}
## R %>%
##   left_join(H, by = "soil_code") %>% 
##   left_join(inv_D, by = "soil_code") %>%
##   left_join(J, by = "soil_code") %>%
##   left_join(dR, by = "soil_code") %>%
##   left_join(dREr, by = "soil_code") %>%
##   left_join(dREa, by = "soil_code") %>%
##   left_join(eDRr, by = "soil_code") %>%
##     left_join(eDRa, by = "soil_code") -> Indices_bact

## names(Indices_bact) <- c("soil_code", "Richness", "Shannon", "Inv_Simpson", "Pielou", 
##                           "Hill_Richness","Hill_Shannon","Hill_Inv_Simpson", 
##                           "Hill_Shannon_evenness", "Hill_Simpson_evenness")
```


***

```{r}
sessionInfo()
rm(list = ls())
```
