---
title: " Oracle soils Burkina Faso t = 0 (#8) MetaG "
author: "Frédéric Mahé"
date: '`r format(Sys.time(), "%d %B %Y")`'

output:
  rmarkdown::html_document:
    theme: lumen
    toc: yes
    toc_float: TRUE
    keep_md: yes
    # code_folding: hide
---

```{r setup, include=FALSE}
rm(list = ls()) # remove all the object before starting
knitr::opts_chunk$set(echo = TRUE)
```

***

#### load required packages

```{r packages, message=FALSE}
library(here)
library(tidyverse)
```


# shotgun metagenomics (SSU rRNA assignment with kraken)

## variables and functions

```{r}
input <- "oracle_metaGt0_79_samples_bracken_SSU_assignment_genuses_rarefied.table3"
taxonomic_levels <- c("domain", "phylum", "class", "order", "family",
                      "genus", "species")  # kingdom == domain
target_taxa <- taxonomic_levels[2:6]
read_passes <- c("R1", "R2")
cbPalette <- c("#56B4E9", "#E69F00", "#009E73", "#F0E442", "#0072B2",
               "#D55E00", "#CC79A7", "#CCCCCC")
seed <- 123
```

```{r}
load_SSU_taxonomic_occurrence_data <- function(filename){
    here::here("data", "MetaG_t0", filename) %>%
        read_tsv(show_col_types = FALSE)
}


remove_level_marker <- function(taxonomic_level) {
    str_remove(taxonomic_level, "^[kpcofgs]__")
}


. %>%
    separate(col = taxonomy, into = taxonomic_levels, sep = "; ") %>%
    mutate(across(.cols = where(is.character),
                  .fns = remove_level_marker)) -> separate_taxonomic_levels


. %>%
    pivot_longer(cols = starts_with("BU-"),
                 names_to = "soil_code",
                 values_to = "reads") %>%
    separate(col = soil_code,
             into = c("soil_code", "read_pass"),
             sep = "_") %>%
    mutate(read_pass = as.factor(read_pass)) -> work_with_R1_and_R2


. %>%
    separate(col = soil_code,
             into = c(NA, "locality", NA),
             sep = "-",
             remove = FALSE) %>%
    mutate(locality = as.factor(locality),
           locality = fct_recode(locality,
                                 Raguitenga = "RT",
                                 Sera = "SER",
                                 Boussouma = "ST",
                                 Tingressene = "TS",
                                 Yilou = "YL")) -> extract_and_recode_locality


lump_and_count <- function(taxonomic_level, target_read_pass) {
    taxonomic_level <- sym(taxonomic_level)

    full_occurrence_table %>%
        filter(domain == "Bacteria" & read_pass == target_read_pass) %>%
        mutate(taxonomic_level = !!taxonomic_level,
               taxonomic_level = replace(taxonomic_level, is.na(taxonomic_level), "Unknown"),
               taxonomic_level = as.factor(taxonomic_level),
               taxonomic_level = fct_lump(taxonomic_level, n = length(cbPalette) - 1, w = reads)) %>%
        count(soil_code, locality, taxonomic_level, wt = reads, name = "reads")
}


plot_taxonomic_barchart <- function(data, legend_title) {
    ggplot(data = data, aes(x = soil_code, y = reads, fill = taxonomic_level)) +
        geom_bar(stat = "identity", position = "fill", colour = "gray", size = 0.2) +
        theme_bw(base_size = 16) +
        scale_fill_manual(name = legend_title, values = cbPalette) +
        scale_y_continuous(labels = scales::percent_format()) +
        ylab("percentage of observed reads") +
        xlab("") +
        facet_wrap(vars(locality), scales = "free_x") +
        theme(axis.text.x  = element_text(angle = 45, vjust = 0.9, hjust = 0.9, size = 10),
              legend.position = c(0.85, 0.25))
}


lump_and_plot <- function(taxonomic_level, read_pass) {
    lump_and_count(taxonomic_level = taxonomic_level,
                   target_read_pass = read_pass) %>%
        plot_taxonomic_barchart(legend_title = taxonomic_level)
}


top_bacterial_family_assignments <- function(target_read_pass) {
    target_read_pass <- sym(target_read_pass)

    full_occurrence_table %>%
        filter(domain == "Bacteria") %>%
        count(family, read_pass, wt = reads, name = "reads") %>%
        pivot_wider(names_from = read_pass, values_from = reads) %>%
        arrange(desc(!!target_read_pass)) %>%
        slice_head(n = 15)
}
```


## load occurrence data

```{r}
load_SSU_taxonomic_occurrence_data(input) %>%
    separate_taxonomic_levels %>%
    work_with_R1_and_R2 %>%
    extract_and_recode_locality -> full_occurrence_table
```


## taxonomic barcharts

### preamble (domain level)

Archaea represent a small fraction of R2, but 17.44% of R1 reads:

```{r}
full_occurrence_table %>%
    count(domain, read_pass, wt = reads, name = "reads")
```

Archaea could play an ecological role:

```{r}
full_occurrence_table %>%
    filter(domain == "Archaea") %>%
    count(family, read_pass, wt = reads, name = "reads")
```

Most Archaea reads are assigned to
[Nitrososphaeraceae](https://en.wikipedia.org/wiki/Nitrososphaera).


### focus on Bacteria

```{r, fig.width=14, fig.height=10}
tibble(taxonomic_level = rep(target_taxa, each = length(read_passes)),
       read_pass = rep(read_passes, times = length(target_taxa))) %>%
    purrr::pmap(., ~ lump_and_plot(taxonomic_level = .x, read_pass = .y))
```


Top bacterial family assignments:

```{r}
read_passes %>%
    purrr::set_names() %>%
    purrr::map(., top_bacterial_family_assignments)
```

Notes regarding family '67-14'. From [Foesel et al.,
2016](https://doi.org/10.1099/ijsem.0.000770) (modified):

> Within the order Solirubrobacterales of the class
> Thermoleophilia. Gram-type-positive, non-spore-forming bacteria,
> some strains were isolated from Namibian semiarid savannah soils and
> are aerobic, chemoheterotrophic mesophiles with broad temperature
> (13-43 °C) and pH (pH 4.5-9.5) ranges for growth.



<!-- ## ecological functions involving nitrogen -->

<!-- Define each ecological group (genus level): -->

<!-- ```{r} -->
<!-- major_legume_nodulating_N_fixers <- c("Allorhizobium-Neorhizobium-Pararhizobium-Rhizobium", -->
<!--                                       "Mesorhizobium", "Sinorhizobium", "Ensifer", -->
<!--                                       "Bradyrhizobium", "Microvirga") -->
<!-- major_actinorhizal_nodulating_N_fixers <- c("Frankia") -->
<!-- AOB_Nitrifiers <- c("Nitrosomonas", "Nitrosospira", "Nitrosococcus") -->
<!-- NOB_Nitrifiers <- c("Nitrobacter", "Nitrospira", "Nitrococcus") -->
<!-- ``` -->


<!-- ### functions -->

<!-- ```{r} -->
<!-- select_ecological_group <- function(ecological_group, group_name) { -->
<!--     full_occurrence_table %>% -->
<!--         filter(genus %in% ecological_group) %>% -->
<!--         count(soil_code, locality, wt = reads, name = "reads") %>% -->
<!--         mutate(group_name = group_name) -->
<!-- } -->
<!-- ``` -->

<!-- if Shapiro-Francia test as a p-value < 0.05, data needs to be transformed -->
<!-- to reach normality. Transformation can be `sqrt()`, `sqrt(sqrt())`, -->
<!-- `log1p()`, `log10()`, `asin(sqrt(x / 100))`. -->

<!-- ```{r} -->
<!-- normality_test <- function(occurrence_table) { -->
<!--     set.seed(seed) -->
<!--     shapiro.test(occurrence_table$reads) -> shapiro_results -->
<!--     print(paste("Shapiro p-value:", shapiro_results$p.value)) -->

<!--     if (shapiro_results$p.value < 0.05) { -->
<!--         print("normality hypothesis is not supported, use a sqrt(x) transformation:") -->
<!--         occurrence_table %>% -->
<!--             mutate(reads = sqrt(reads)) -> occurrence_table -->
<!--     } else { -->
<!--         print("normality hypothesis cannot be rejected, no transformation needed.") -->
<!--     } -->

<!--     occurrence_table -->
<!-- } -->
<!-- ``` -->


<!-- ### major legume nodulating N-fixers -->

<!-- see [Shamseldin et al. (2017)](https://doi.org/10.1007/s13199-016-0462-3): -->

<!-- ```{r} -->
<!-- select_ecological_group(ecological_group = major_legume_nodulating_N_fixers, -->
<!--                         group_name = "legume nodulating") -> legume_nodulating -->
<!-- ``` -->


<!-- ##### normality test (Shapiro-Francia test) -->

<!-- ```{r} -->
<!-- normality_test(occurrence_table = legume_nodulating) -> legume_nodulating -->
<!-- hist(legume_nodulating$reads) -->
<!-- print("Quantile-Quantile plot") -->
<!-- ggpubr::ggqqplot(legume_nodulating$reads, ylab = "Number of reads?") -->
<!-- ``` -->


<!-- ### major actinorhizal nodulating N-fixers -->

<!-- *Frankia* is the only known genus in Africa. -->

<!-- ```{r} -->
<!-- select_ecological_group(ecological_group = major_actinorhizal_nodulating_N_fixers, -->
<!--                         group_name = "actinorhizal nodulating") -> actinorhizal_nodulating -->
<!-- ``` -->


<!-- ##### normality test (Shapiro-Francia test) -->

<!-- ```{r} -->
<!-- normality_test(occurrence_table = actinorhizal_nodulating) -> actinorhizal_nodulating -->
<!-- hist(actinorhizal_nodulating$reads) -->
<!-- print("Quantile-Quantile plot") -->
<!-- ggpubr::ggqqplot(actinorhizal_nodulating$reads, ylab = "Number of reads?") -->
<!-- ``` -->


<!-- ### ammonium oxidizing bacteria (AOB Nitrifiers) -->

<!-- ```{r} -->
<!-- select_ecological_group(ecological_group = AOB_Nitrifiers, -->
<!--                         group_name = "AOB Nitrifiers") -> AOB -->
<!-- ``` -->


<!-- ##### normality test (Shapiro-Francia test) -->

<!-- ```{r} -->
<!-- normality_test(occurrence_table = AOB) -> AOB -->
<!-- hist(AOB$reads) -->
<!-- print("Quantile-Quantile plot") -->
<!-- ggpubr::ggqqplot(AOB$reads, ylab = "Number of reads?") -->
<!-- ``` -->


<!-- ### nitrite oxidizing bacteria (NOB Nitrifiers) -->

<!-- ```{r} -->
<!-- select_ecological_group(ecological_group = NOB_Nitrifiers, -->
<!--                         group_name = "NOB Nitrifiers") -> NOB -->
<!-- ``` -->


<!-- ##### normality test (Shapiro-Francia test) -->

<!-- ```{r} -->
<!-- normality_test(occurrence_table = NOB) -> NOB -->
<!-- hist(NOB$reads) -->
<!-- print("Quantile-Quantile plot") -->
<!-- ggpubr::ggqqplot(NOB$reads, ylab = "Number of reads?") -->
<!-- ``` -->


<!-- ### locality effect -->

<!-- #### functions -->

<!-- ```{r} -->
<!-- make_kruskal_test <- function(ecological_group, group_name) { -->
<!--     kruskal.test(formula = reads ~ locality, -->
<!--                  data = ecological_group, -->
<!--                  na.action = na.omit) -> kruskal_results -->

<!--     kruskal_results$p.value %>% -->
<!--         round(digits = 4) %>% -->
<!--         paste("Kruskal-Wallis rank sum test", -->
<!--               "\nLocality P = ", ., sep = "") -> local_message -->

<!--     tibble(ecological_group = ecological_group, -->
<!--            message = local_message, -->
<!--            group_name = group_name) -->
<!-- } -->
<!-- ``` -->

<!-- ```{r} -->
<!-- make_dunn_test <- function(ecological_group, group_name) { -->
<!--     dunn.test::dunn.test(ecological_group$reads, -->
<!--                          ecological_group$locality, -->
<!--                          method = "bh") -> dunn_results -->

<!--     tibble(comparisons = dunn_results$comparisons, -->
<!--            p_adjusted = dunn_results$P.adjusted, -->
<!--            group_name = group_name) -->
<!-- } -->
<!-- ``` -->

<!-- #### kruskall Wallis test (two groups) -->


<!-- ```{r} -->
<!-- bind_rows( -->
<!--     make_kruskal_test(ecological_group = legume_nodulating, -->
<!--                       group_name = "legume nodulating"), -->
<!--     make_kruskal_test(ecological_group = actinorhizal_nodulating, -->
<!--                       group_name = "actinorhizal_nodulating"), -->
<!--     make_kruskal_test(ecological_group = AOB, -->
<!--                       group_name = "AOB Nitrifiers"), -->
<!--     make_kruskal_test(ecological_group = NOB, -->
<!--                       group_name = "NOB Nitrifiers") -->
<!-- ) %>% -->
<!--     group_by(group_name, message) %>% -->
<!--     summarize(ymax = 1.1 * max(ecological_group$reads)) %>% -->
<!--     ungroup() %>% -->
<!--     mutate(group_name = str_replace(group_name, "_", " ")) -> kruskal_tests -->
<!-- ``` -->

<!-- note that the tibble is simplified (only one p-value per ecological -->
<!-- group), to facilitate the inclusion in facetted boxplots. -->


<!-- #### Dunn test (multi groups) -->

<!-- ```{r} -->
<!-- bind_rows( -->
<!--     make_dunn_test(ecological_group = legume_nodulating, -->
<!--                    group_name = "legume nodulating"), -->
<!--     make_dunn_test(ecological_group = actinorhizal_nodulating, -->
<!--                    group_name = "actinorhizal_nodulating"), -->
<!--     make_dunn_test(ecological_group = AOB, -->
<!--                    group_name = "AOB Nitrifiers"), -->
<!--     make_dunn_test(ecological_group = NOB, -->
<!--                    group_name = "NOB Nitrifiers") -->
<!-- ) -> dunn_tests -->
<!-- ``` -->

<!-- These test results are used by Hervé in the boxplots. -->


<!-- #### boxplots -->

<!-- ```{r} -->
<!-- bind_rows( -->
<!--     legume_nodulating, -->
<!--     actinorhizal_nodulating, -->
<!--     AOB, -->
<!--     NOB -->
<!-- ) -> occurrences -->
<!-- ``` -->

<!-- by default, localities are plotted in alphabetical order, we need to -->
<!-- know which one is the first: -->

<!-- ```{r} -->
<!-- occurrences %>% -->
<!--     distinct(locality) %>% -->
<!--     arrange(locality) %>% -->
<!--     head(n = 1) %>% -->
<!--     pull() -> first_locality -->
<!-- ``` -->


<!-- ```{r} -->
<!-- occurrences %>% -->
<!--     ggplot(aes(x = locality, y = reads, fill = locality)) + -->
<!--     geom_boxplot(outlier.shape = NA) + -->
<!--     scale_fill_manual(values = cbPalette, guide = "none") + -->
<!--     geom_jitter(position = position_jitter(0.1), -->
<!--                 alpha = 1/5, cex = 2.5) + -->
<!--     geom_text(data = kruskal_tests, -->
<!--               aes(x = first_locality, y = ymax, label = message), -->
<!--               inherit.aes = FALSE, -->
<!--               hjust = 0.1, vjust = 1) + -->
<!--     facet_wrap(vars(group_name), scales = "free_y") + -->
<!--     xlab("") + -->
<!--     theme_bw(base_size = 16) -->
<!-- ``` -->


<!-- #### clean up -->

<!-- ```{r} -->
<!-- rm(full_occurrence_table, legume_nodulating, actinorhizal_nodulating, AOB, NOB, first_locality) -->
<!-- ``` -->

<!-- roadmap: -->

<!-- - test git pull, -->
<!-- - ask Hervé to review the Nitrogen section, -->
<!-- - NA should be marked as unknown?, -->
<!-- - add '(Frankia)' to the legend 'Actinorhizal N-Fixer', as there is -->
<!--   only one genus with that function -->

***

```{r}
sessionInfo()
rm(list = ls())
```
