---
title: " Oracle soils Burkina Faso t = 0 (#8) "
author: "Frédéric Mahé"
date: '`r format(Sys.time(), "%d %B %Y")`'

output:
  rmarkdown::html_document:
    theme: lumen
    toc: yes
    toc_float: TRUE
    keep_md: yes
    # code_folding: hide
---

```{r setup, include=FALSE}
rm(list = ls()) # remove all the object before starting
knitr::opts_chunk$set(echo = TRUE)
```

***

#### load required packages

```{r packages, message=FALSE}
library(here)
library(tidyverse)
```


# SSU rRNA metabarcoding (16S V3-V4)

## variables and functions

```{r}
input <- "Oracle_soils_16S_341F_785R_87_samples.OTU.filtered.cleaved.nosubstringOTUs.mumu.table4"
cbPalette <- c("#56B4E9", "#E69F00", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#CCCCCC")
seed <- 123
```

```{r}
load_16S_taxonomic_occurrence_data <- function(filename){
    here::here("data", "MetaB_t0", filename) %>%
        read_tsv(show_col_types = FALSE)
}


lump_and_count <- function(taxonomic_level) {
    taxonomic_level <- sym(taxonomic_level)

    full_occurrence_table %>%
        filter(domain == "Bacteria") %>%
        mutate(taxonomic_level = !!taxonomic_level) %>%
        mutate(taxonomic_level = replace(taxonomic_level, is.na(taxonomic_level), "Unknown")) %>%
        mutate(taxonomic_level = fct_lump(taxonomic_level, n = length(cbPalette) - 1, w = reads)) %>%
        count(soil_code, locality, taxonomic_level, wt = reads, name = "reads")
}


plot_taxonomic_barchart <- function(data, legend_title) {
    ggplot(data = data, aes(x = soil_code, y = reads, fill = taxonomic_level)) +
        geom_bar(stat = "identity", position = "fill", colour = "gray", size = 0.2) +
        theme_bw(base_size = 16) +
        scale_fill_manual(name = legend_title, values = cbPalette) +
        scale_y_continuous(labels = scales::percent_format()) +
        ylab("percentage of observed reads") +
        xlab("") +
        facet_wrap(vars(locality), scales = "free_x") +
        theme(axis.text.x  = element_text(angle = 45, vjust = 0.9, hjust = 0.9, size = 10),
              legend.position = c(0.85, 0.25))
}


lump_and_plot <- function(taxonomic_level) {
    lump_and_count(taxonomic_level = taxonomic_level) %>%
        plot_taxonomic_barchart(legend_title = taxonomic_level)
}
```


## load occurrence data

```{r}
load_16S_taxonomic_occurrence_data(input) -> full_occurrence_table
```


## taxonomic barcharts

### preamble (domain level)

Archaea only represent a small fraction:

```{r}
full_occurrence_table %>%
    count(domain, wt = reads, sort = TRUE, name = "reads")
```

However, Archaea could play an ecological role:

```{r}
full_occurrence_table %>%
    filter(domain == "Archaea") %>%
    count(family, wt = reads, sort = TRUE, name = "reads")
```


### focus on Bacteria

```{r, fig.width=14, fig.height=10}
lump_and_plot(taxonomic_level = "phylum")
lump_and_plot(taxonomic_level = "class")
lump_and_plot(taxonomic_level = "order")
lump_and_plot(taxonomic_level = "family")
lump_and_plot(taxonomic_level = "genus")
```


Top bacterial family assignments:

```{r}
full_occurrence_table %>%
    filter(domain == "Bacteria") %>%
    count(family, wt = reads, sort = TRUE, name = "reads") %>%
    slice_head(n = 15)
```

Notes regarding family '67-14'. From [Foesel et al.,
2016](https://doi.org/10.1099/ijsem.0.000770) (modified):

> Within the order Solirubrobacterales of the class
> Thermoleophilia. Gram-type-positive, non-spore-forming bacteria,
> some strains were isolated from Namibian semiarid savannah soils and
> are aerobic, chemoheterotrophic mesophiles with broad temperature
> (13-43 °C) and pH (pH 4.5-9.5) ranges for growth.



## ecological functions involving nitrogen

Define each ecological group (genus level):

```{r}
major_legume_nodulating_N_fixers <- c("Allorhizobium-Neorhizobium-Pararhizobium-Rhizobium",
                                      "Mesorhizobium", "Sinorhizobium", "Ensifer",
                                      "Bradyrhizobium", "Microvirga")
major_actinorhizal_nodulating_N_fixers <- c("Frankia")
AOB_Nitrifiers <- c("Nitrosomonas", "Nitrosospira", "Nitrosococcus")
NOB_Nitrifiers <- c("Nitrobacter", "Nitrospira", "Nitrococcus")
```


### functions

```{r}
select_ecological_group <- function(ecological_group, group_name) {
    full_occurrence_table %>%
        filter(genus %in% ecological_group) %>%
        count(soil_code, locality, wt = reads, name = "reads") %>%
        mutate(group_name = group_name)
}
```

if Shapiro-Francia test as a p-value < 0.05, data needs to be transformed
to reach normality. Transformation can be `sqrt()`, `sqrt(sqrt())`,
`log1p()`, `log10()`, `asin(sqrt(x / 100))`.

```{r}
normality_test <- function(occurrence_table) {
    set.seed(seed)
    shapiro.test(occurrence_table$reads) -> shapiro_results
    print(paste("Shapiro p-value:", shapiro_results$p.value))

    if (shapiro_results$p.value < 0.05) {
        print("normality hypothesis is not supported, use a sqrt(x) transformation:")
        occurrence_table %>%
            mutate(reads = sqrt(reads)) -> occurrence_table
    } else {
        print("normality hypothesis cannot be rejected, no transformation needed.")
    }

    occurrence_table
}
```


### major legume nodulating N-fixers

see [Shamseldin et al. (2017)](https://doi.org/10.1007/s13199-016-0462-3):

```{r}
select_ecological_group(ecological_group = major_legume_nodulating_N_fixers,
                        group_name = "legume nodulating") -> legume_nodulating
```


##### normality test (Shapiro-Francia test)

```{r}
normality_test(occurrence_table = legume_nodulating) -> legume_nodulating
hist(legume_nodulating$reads)
print("Quantile-Quantile plot")
ggpubr::ggqqplot(legume_nodulating$reads, ylab = "Number of reads?")
```


### major actinorhizal nodulating N-fixers

*Frankia* is the only known genus in Africa.

```{r}
select_ecological_group(ecological_group = major_actinorhizal_nodulating_N_fixers,
                        group_name = "actinorhizal nodulating") -> actinorhizal_nodulating
```


##### normality test (Shapiro-Francia test)

```{r}
normality_test(occurrence_table = actinorhizal_nodulating) -> actinorhizal_nodulating
hist(actinorhizal_nodulating$reads)
print("Quantile-Quantile plot")
ggpubr::ggqqplot(actinorhizal_nodulating$reads, ylab = "Number of reads?")
```


### ammonium oxidizing bacteria (AOB Nitrifiers)

```{r}
select_ecological_group(ecological_group = AOB_Nitrifiers,
                        group_name = "AOB Nitrifiers") -> AOB
```


##### normality test (Shapiro-Francia test)

```{r}
normality_test(occurrence_table = AOB) -> AOB
hist(AOB$reads)
print("Quantile-Quantile plot")
ggpubr::ggqqplot(AOB$reads, ylab = "Number of reads?")
```


### nitrite oxidizing bacteria (NOB Nitrifiers)

```{r}
select_ecological_group(ecological_group = NOB_Nitrifiers,
                        group_name = "NOB Nitrifiers") -> NOB
```


##### normality test (Shapiro-Francia test)

```{r}
normality_test(occurrence_table = NOB) -> NOB
hist(NOB$reads)
print("Quantile-Quantile plot")
ggpubr::ggqqplot(NOB$reads, ylab = "Number of reads?")
```


### locality effect

#### functions

```{r}
make_kruskal_test <- function(ecological_group, group_name) {
    kruskal.test(formula = reads ~ locality,
                 data = ecological_group,
                 na.action = na.omit) -> kruskal_results

    kruskal_results$p.value %>%
        round(digits = 4) %>%
        paste("Kruskal-Wallis rank sum test",
              "\nLocality P = ", ., sep = "") -> local_message

    tibble(ecological_group = ecological_group,
           message = local_message,
           group_name = group_name)
}
```

```{r}
make_dunn_test <- function(ecological_group, group_name) {
    dunn.test::dunn.test(ecological_group$reads,
                         ecological_group$locality,
                         method = "bh") -> dunn_results

    tibble(comparisons = dunn_results$comparisons,
           p_adjusted = dunn_results$P.adjusted,
           group_name = group_name)
}
```

#### kruskall Wallis test (two groups)


```{r}
bind_rows(
    make_kruskal_test(ecological_group = legume_nodulating,
                      group_name = "legume nodulating"),
    make_kruskal_test(ecological_group = actinorhizal_nodulating,
                      group_name = "actinorhizal_nodulating"),
    make_kruskal_test(ecological_group = AOB,
                      group_name = "AOB Nitrifiers"),
    make_kruskal_test(ecological_group = NOB,
                      group_name = "NOB Nitrifiers")
) %>%
    group_by(group_name, message) %>%
    summarize(ymax = 1.1 * max(ecological_group$reads)) %>%
    ungroup() %>%
    mutate(group_name = str_replace(group_name, "_", " ")) -> kruskal_tests
```

note that the tibble is simplified (only one p-value per ecological
group), to facilitate the inclusion in facetted boxplots.


#### Dunn test (multi groups)

```{r}
bind_rows(
    make_dunn_test(ecological_group = legume_nodulating,
                   group_name = "legume nodulating"),
    make_dunn_test(ecological_group = actinorhizal_nodulating,
                   group_name = "actinorhizal_nodulating"),
    make_dunn_test(ecological_group = AOB,
                   group_name = "AOB Nitrifiers"),
    make_dunn_test(ecological_group = NOB,
                   group_name = "NOB Nitrifiers")
) -> dunn_tests
```

These test results are used by Hervé in the boxplots.


#### boxplots

```{r}
bind_rows(
    legume_nodulating,
    actinorhizal_nodulating,
    AOB,
    NOB
) -> occurrences
```

by default, localities are plotted in alphabetical order, we need to
know which one is the first:

```{r}
occurrences %>%
    distinct(locality) %>%
    arrange(locality) %>%
    head(n = 1) %>%
    pull() -> first_locality
```


```{r}
occurrences %>%
    ggplot(aes(x = locality, y = reads, fill = locality)) +
    geom_boxplot(outlier.shape = NA) +
    scale_fill_manual(values = cbPalette, guide = "none") +
    geom_jitter(position = position_jitter(0.1),
                alpha = 1/5, cex = 2.5) +
    geom_text(data = kruskal_tests,
              aes(x = first_locality, y = ymax, label = message),
              inherit.aes = FALSE,
              hjust = 0.1, vjust = 1) +
    facet_wrap(vars(group_name), scales = "free_y") +
    xlab("") +
    theme_bw(base_size = 16)
```


#### clean up

```{r}
rm(full_occurrence_table, legume_nodulating, actinorhizal_nodulating, AOB, NOB, first_locality)
```

roadmap:

- ask Hervé to review the Nitrogen section,
- NA should be marked as unknown?,
- add '(Frankia)' to the legend 'Actinorhizal N-Fixer', as there is
  only one genus with that function

***

```{r}
sessionInfo()
rm(list = ls())
```
